---
title: "Introduction to scr3way"
author: "Diogo Ribeiro"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to scr3way}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
library(scr3way)
```

# Overview

`scr3way` implements simultaneous clustering and dimensionality reduction for
three-way data. The package is built around the Simultaneous Clustering and
Reduction (SCR) methodology of Rocci, Vichi, and Ranalli (2025).

Suppose that each observation contains (J) variables measured on (K)
occasions. After vectorization, each observation has dimension (JK), but the
three-way structure is retained through separate variable and occasion modes.

The package distinguishes a validated baseline from research extensions.

# Baseline models

The baseline comparison contains three models.

- **S3** is the three-way SCR model. Its component means use a Tucker2
  discriminating subspace and its common covariance is separable.
- **S2** applies simultaneous clustering and reduction after vectorizing the
  observations.
- **H** is the ordinary homoscedastic Gaussian mixture comparator.

The main fitting functions are `fit_scr_s3()`, `fit_scr_s2()`, and
`fit_homoscedastic_gaussian_mixture()`.

# Tucker3 centroid reduction

The Tucker3 extension also reduces the group or centroid mode. For ranks
((P,Q,R)), the component means are represented through centroid, variable,
and occasion bases together with a Tucker core.

A small deterministic fit can be created as follows.

```{r tucker3-fit}
set.seed(123)

n <- 40L
groups <- 2L
variables <- 2L
occasions <- 2L

labels <- rep(seq_len(groups), each = n / groups)
means <- rbind(
  c(-2, 0, -1, 0),
  c(2, 0, 1, 0)
)

X <- means[labels, , drop = FALSE] +
  matrix(rnorm(n * variables * occasions, sd = 0.4), nrow = n)

membership <- matrix(0.1, nrow = n, ncol = groups)
membership[cbind(seq_len(n), labels)] <- 0.9
membership <- membership / rowSums(membership)

fit <- fit_scr_s3_tucker3(
  X = X,
  membership = membership,
  centroid_rank = 1,
  variable_rank = 1,
  occasion_rank = 1,
  variable_covariance = diag(variables),
  occasion_covariance = diag(occasions),
  max_iter = 10,
  inner_max_iter = 10
)

fit$ranks
fit$bic
```

# Structural model selection

For fixed (G), `select_scr_tucker3_model()` compares candidate
((P,Q,R)) structures. The package supports BIC and ICL. An outer layer,
`select_scr_tucker3_groups()`, extends the search to candidate values of
(G).

The package uses the convention

[
\mathrm{BIC} = 2\ell - k\log n,
]

so larger values are preferred.

ICL subtracts a classification-entropy penalty,

[
\mathrm{ICL} = \mathrm{BIC} - 2H(U).
]

Subsampling stability is available separately through
`scr_tucker3_stability()` and `select_scr_tucker3_stable()`.

# Covariance extensions

The S3 baseline assumes exact separability,

[
\Sigma = \Sigma_O \otimes \Sigma_V.
]

The package provides tools for measuring departures from separability and for
studying the controlled nugget family

[
\Sigma(\tau)
=
\Sigma_O \otimes \Sigma_V + \tau I.
]

Relevant functions include `scr_nearest_kronecker_covariance()`,
`scr_kronecker_nugget_covariance()`, `profile_scr_tucker3_nugget()`, and
`refine_scr_tucker3_nugget()`.

# Sparse discriminating subspaces

Sparse research utilities apply row-group thresholding to the variable and
occasion loading matrices while restoring covariance-metric orthonormality.

The main sparse functions are `scr_sparse_metric_basis()`,
`scr_sparse_tucker3_projection()`, `scr_sparse_tucker3_path()`, and
`select_scr_sparse_tucker3()`.

These functions currently implement post-fit projection and conditional
support selection rather than a fully penalized mixture estimator.

# Robust Student-t likelihood

The robust layer evaluates a multivariate Student-t mixture under the Tucker3
mean structure. For squared Mahalanobis distance (delta_{ig}), the latent
precision weight is

[
w_{ig} = \frac{\nu + JK}{\nu + \delta_{ig}}.
]

A remote observation therefore receives less effective weight.

```{r student-t}
robust <- scr_tucker3_student_loglik(
  X = X,
  means = fit$M,
  probabilities = fit$probabilities,
  variable_scale = fit$SV,
  occasion_scale = fit$SO,
  degrees_freedom = 6
)

head(robust$observation_weights)
```

# Reproducibility and research status

The package keeps the published SCR baseline conceptually separate from
research extensions. The repository also contains development-only regression
material used to compare the R implementation with the authors' public MATLAB
reference implementation.

For the detailed reconstruction history, methodological assumptions, and
research roadmap, see the repository document
`docs/scr-research-program.md`.
